Exploratory Analysis Script

### 2024-04-05 code/analysis-code/introanalysis-v1.qmd
Project Part 4
Completed full analysis and updated all files.

---
title: "Statistical Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

@@

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling
#rm(list=ls()) #clear the environment
```

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear_processed.rds")
#load data
ML_linear <- readRDS(data_location)
```

Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry.

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm1 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = ML_linear)

# Summarize model fit with tidiers
ML_lm1_summary <- summary(ML_lm1) # Summary of fit
tidy(ML_lm1, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_lm1) # Model-level statistics

# Produce predictions for RMSE and R-squared
ML_lm1_pred <- predict(ML_lm1, new_data = ML_linear %>% select(-Weightgrams))
ML_lm1_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_lm1_pred <- bind_cols(ML_lm1_pred, ML_linear %>% select(Weightgrams))
ML_lm1_pred

# Plot observed versus predicted
plot1 <- ggplot(ML_lm1_pred, aes(x = Weightgrams, y = .pred)) +
  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
  geom_point(alpha = 0.5) +
  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
  coord_obs_pred()
plot1

# Create metric set including RMSE and R-squared
metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
metrics_lm1(ML_lm1_pred, truth = Weightgrams, estimate = .pred)

# Create a gtsummary table directly from the model
table2 <- tbl_regression(ML_lm1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table2

# Save regression output into table
saveRDS(table2, file = here::here("results", "tables", "table2_final.rds"))
```
Now let's run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, we can use a scatterplot.

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm2 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Age, data = ML_linear)

# Summarize model fit with tidiers
ML_lm2_summary <- summary(ML_lm2) # Summary of fit
tidy(ML_lm2, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_lm2) # Model-level statistics

# Plot data
plot2 <- ggplot(ML_linear, aes(x = Age, y = Weightgrams)) +
  geom_point(alpha = 0.5) +
  labs(x = "Maternal Age", y = "Infant Birth Weight (grams)") + #Scale and size x- and y-axis uniformly
  geom_smooth(method = "lm", se = FALSE) # Add a linear trend line
  theme_minimal() #set plot theme
plot2

# Produce predictions for RMSE and R-squared
ML_lm2_pred <- predict(ML_lm2, new_data = ML_linear %>% select(-Weightgrams))
ML_lm2_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_lm2_pred <- bind_cols(ML_lm2_pred, ML_linear %>% select(Weightgrams))
ML_lm2_pred

# Create metric set including RMSE and R-squared
metrics_lm1 <- metric_set(rmse, rsq, mae) # MAE for fun
metrics_lm1(ML_lm2_pred, truth = Weightgrams, estimate = .pred)

# Create a gtsummary table directly from the model
table3 <- tbl_regression(ML_lm2, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table3

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3_final.rds"))
```

Now let's run a slightly more complicated model that includes maternal age, nationality, and gestational status.

@@

# Create a gtsummary table directly from the model
table4 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table4

# Save regression output into table
saveRDS(table4, file = here::here("results", "tables", "table4_final.rds"))
```

@@

>>>>

---
title: "Intro Statistical Analysis Script"
author: "Cassia Roth"
date: "`r Sys.Date()`"
output: html_document
---

@@

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling

#rm(list=ls()) #clear the environment
```

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

Let's run a simple regression with the outcome variable (birth weight in grams) and the exposure variable maternal ancestry. I will use the recipe and workflow features of `tidymodels`.

```{r}
# Model using recipes and workflows
# Define recipe for model
ML_recipe1 <- recipe(Weightgrams ~ ModifiedColor, data = ML_linear) %>% 
  step_relevel(ModifiedColor, ref_level = "Euro-Descent") %>% # Set reference level for ModifiedColor
  step_dummy(ModifiedColor) # Convert ModifiedColor to dummy variable

# Specify model
ML_spec1 <- linear_reg() %>% 
  set_engine("lm")

# Create work flow
ML_wf1 <- workflow() %>%
  add_recipe(ML_recipe1) %>%
  add_model(ML_spec1)

# Fit the work flow
ML_fit1 <- fit(ML_wf1, data = ML_linear)

# Summarize model fit with tidiers
ML_summary1 <- summary(ML_fit1) # Summary of fit
tidy(ML_fit1, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit1) # Model-level statistics

# Produce predictions for R-squared
ML_fit1_pred <- predict(ML_fit1, new_data = ML_linear %>% select(-Weightgrams))
ML_fit1_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_fit1_pred <- bind_cols(ML_fit1_pred, ML_linear %>% select(Weightgrams))
ML_fit1_pred

# Plot observed versus predicted
plot1 <- ggplot(ML_fit1_pred, aes(x = Weightgrams, y = .pred)) +
  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
  geom_point(alpha = 0.5) +
  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
  coord_obs_pred()
plot1

# Create metric set including RMSE and R-squared
metrics_fit1 <- metric_set(rmse, rsq)
metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)

# Convert the output to a data frame and mutate the .estimate column to numeric
metrics_fit1 <- as.data.frame(metrics_fit1(ML_fit1_pred, truth = Weightgrams, estimate = .pred)) %>%
  mutate(.estimate = as.numeric(.estimate))
options(scipen = 999) # set opens to avoid scientific notation
metrics_fit1

# Create a gtsummary table directly from the model
table2 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))

#Print table
table2

# Save regression output into table
saveRDS(table2, file = here::here("results", "tables", "table2.rds"))
```
Now let's run a second simple bivariate analysis, looking at infant birth weight as a function of maternal age. This time, when plotting weight as a function of maternal age, we can use a scatterplot.

```{r}
# Model using recipes and workflows
# Create recipe
ML_recipe_age <- recipe(Weightgrams ~ Age, data = ML_linear)

# Specify model
ML_spec_age <- linear_reg() %>% 
  set_engine("lm")

# Create work flow
ML_wf_age <- workflow() %>%
  add_recipe(ML_recipe_age) %>%
  add_model(ML_spec_age)

# Fit the work flow
ML_fit_age <- fit(ML_wf_age, data = ML_linear)

# Summarize model fit with tidiers
ML_summary_age <- summary(ML_fit_age) # Summary of fit
tidy(ML_fit_age, conf.int = TRUE) # Coefficients and confidence intervals
glance(ML_fit_age) # Model-level statistics

# Produce predictions for RMSE and R-squared
ML_age_pred <- predict(ML_fit_age, new_data = ML_linear %>% select(-Weightgrams))
ML_age_pred

# Predicted numeric outcome named .pred. Now we will match predicted values with corresponding observed outcome values
ML_age_pred <- bind_cols(ML_age_pred, ML_linear %>% select(Weightgrams))
ML_age_pred

# Plot observed versus predicted
plot_age <- ggplot(ML_age_pred, aes(x = Weightgrams, y = .pred)) +
  geom_abline(lty = 2) + # Add a dashed line to represent the 1:1 line
  geom_point(alpha = 0.5) +
  labs(x = "Observed", y = "Predicted") + #Scale and size x- and y-axis uniformly
  coord_obs_pred()
plot_age

# Create metric set including RMSE and R-squared
metrics_fit_age <- metric_set(rmse, rsq)
metrics_fit_age(ML_age_pred, truth = Weightgrams, estimate = .pred)

# Convert the output to a data frame and mutate the .estimate column to numeric
metrics_fit_age <- as.data.frame(metrics_fit_age(ML_age_pred, truth = Weightgrams, estimate = .pred)) %>%
  mutate(.estimate = as.numeric(.estimate))
options(scipen = 999) # set opens to avoid scientific notation
metrics_fit_age

# Create a gtsummary table directly from the model
table_age <-
  tbl_regression(ML_fit_age, exponentiate = FALSE, intercept = TRUE, show_single_row = everything()) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))

#Print table
table_age

# Save regression output into table
saveRDS(table_age, file = here::here("results", "tables", "tableage_final.rds"))
```

Now let's run a slightly more complicated model that includes maternal skin color, age, nationality, and gestational status.

@@

# Create a gtsummary table directly from the model
table3 <- tbl_regression(ML_lm3, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table3

# Save regression output into table
saveRDS(table3, file = here::here("results", "tables", "table3.rds"))
```

@@

Now let's try to run a model where skin color is differentiated according to the original data, or with three categories (White, Black, and Mixed Race). We'll start with a simple model of infant birth weight as a function of maternal skin color. Then we will run a model with all the predictors.

```{r}
# Simple linear model
# Set reference level for Color
ML_linear$Color <- relevel(ML_linear$Color, ref = "White")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm4 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color, data = ML_linear)

# Summarize model fit with tidiers
ML_lm4_summary <- summary(ML_lm4) # Summary of fit
tidy(ML_lm4, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm4) # Model-level statistics

# Create a gtsummary table directly from the model
table4 <- tbl_regression(ML_lm4, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table4

# Save regression output into table
saveRDS(table4, file = here::here("results", "tables", "table4.rds"))

# Complex linear model
# Set reference level for Color
ML_linear$Color <- relevel(ML_linear$Color, ref = "White")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm5 <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ Color + Age + ModifiedStatus + ModifiedNationality, data = ML_linear)

# Summarize model fit with tidiers
ML_lm5_summary <- summary(ML_lm5) # Summary of fit
tidy(ML_lm5, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
glance(ML_lm5) # Model-level statistics

# Create a gtsummary table directly from the model
table5 <- tbl_regression(ML_lm5, exponentiate = FALSE, intercept = TRUE) %>% 
  bold_p() %>% 
  add_glance_table(include = c(r.squared, adj.r.squared, nobs)) %>%
  modify_column_unhide(columns = c(statistic, std.error))
table5

# Save regression output into table
saveRDS(table5, file = here::here("results", "tables", "table5.rds"))
```

### 2024-04-05 code/eda-code/eda-v1.qmd
Project Part 4
Completed full analysis and updated all files.

# Create dummy variables for status and change to English
ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")

@@

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary_final.rds"))
```

@@

```{r}
# Create summary
summary = skimr::skim(ML_linear)
print(summary)

# save to file
lineartable_file = here("results", "tables", "lineartable.rds")
saveRDS(summary, file = lineartable_file)
```

@@

# Create dummy variables using recode factor from dplyr
 ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")

# Create dummy variables for status and change to English
ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")

@@

# Change observations from Portuguese to English
ML_linear <- ML_linear %>%
  mutate(
    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )

# Save new summary data with recoded factor variables
saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear_final.rds"))
```

>>>>

# Create dummy variables for status and change to English
ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multiparous", "Multipara" = "Multiparous", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")

@@

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
```

@@

Now I will create dummy variables.

@@

# Create dummy variables using recode factor from dplyr
ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")

# Create dummy variables for status and change to English
ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Nullipara", "Nulipara" = "Nullipara")

@@

# Change observations from Portuguese to English
ML_linear <- ML_linear %>%
  mutate(
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )

str(ML_linear)

# Save new summary data with recoded factor variables
saveRDS(ML_linear, file = here("data", "processed-data", "ML_linear.rds"))
```
